---
title: "archetpe_analysis_multiome"
author: "Sarah Groves"
date: "10/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data and packages

reference SCLC multiome analysis.Rmd for in depth description of sclc RDS.
```{r setup-libraries, include=FALSE}
library(Signac)
library(Seurat)
library(GenomeInfoDb)
library(EnsDb.Mmusculus.v79) 
library(ggplot2)
library(patchwork)
library(dplyr)
# install.packages('devtools')
library(devtools)
# BiocManager::install("scater")
library(scater)

# Load plot_ly
library(plotly)

# install loomR from GitHub using the remotes package 
# remotes::install_github(repo ='mojaveazure/loomR', ref = 'develop')
set.seed(1234)
```

```{r, import_pareto, include=FALSE}
# install.packages("BiocManager") # for installing BioConductor dependencies
# BiocManager::install("vitkl/ParetoTI", dependencies = c("Depends", "Imports", "LinkingTo"))
# ParetoTI::install_py_pcha(method = "conda",
                          # extra_packages = c("tensorflow", "tensorflow-probability",
                          #               "pandas", "keras", "h5py",
                          #               "geosketch", "pydot", "scikit-learn==0.20",
                          #               "umap-learn"))
# ParetoTI::install_py_pcha()
library(reticulate)
reticulate::use_condaenv("reticulate_PCHA", conda = "auto",
                         required = TRUE) # set TRUE to force R to use reticulate_PCHA
reticulate::py_discover_config("py_pcha")

library(ParetoTI)
```

Using data from Debbie, with her PCA and labels.
```{r read-data}

<<<<<<< Updated upstream:archetype-plasticity-notebooks/M1/multiome-archetypes.Rmd
sclc <- readRDS('../data/TKO_10_12.rds')
sclc
```
=======
It looks like ciliated cells are dominating PC2, so we are going to remove those cells and just look at the four archetypes defining the other cell types. 
```{r without-ciliated, message=FALSE}
TKO <- readRDS('../data/TKO_with_MAGIC_11_1.rds')
DefaultAssay(TKO) <- "Imputed_counts"
>>>>>>> Stashed changes:notebooks/multiome-archetypes MAGIC imputed.Rmd

```{r, include = FALSE}
# write.csv(sclc@assays$SCT@scale.data, '../data/data_from_debbie/scaled-SCT-data.csv')
# write.csv(sclc@reductions$pca@cell.embeddings, '../data/data_from_debbie/pca-embedding.csv')
# write.csv(sclc@reductions$pca@feature.loadings, '../data/data_from_debbie/pca-feature-loadings.csv')
# 
# sclc <- ProjectDim(sclc, reduction = 'pca')
# 
# write.csv(sclc@reductions$pca@feature.loadings.projected, '../data/data_from_debbie/pca-feature-loadings-projected.csv')
```

<<<<<<< Updated upstream:archetype-plasticity-notebooks/M1/multiome-archetypes.Rmd
# Plot UMAP of data for reference
```{r extract-umap, include = FALSE}
# Extract tSNE information from Seurat Object
umap_1 <- sclc[["umap"]]@cell.embeddings[,1]
umap_2 <- sclc[["umap"]]@cell.embeddings[,2]
=======
TKO_pca <- TKO@reductions$pca@cell.embeddings
TKO_pca <- t(TKO_pca)
TKO_pca <- TKO_pca[1:20,] #keep only top 18 PCs
loadings<- TKO@reductions$pca@feature.loadings.projected
loadings <- as.matrix(loadings)
ElbowPlot(TKO)
>>>>>>> Stashed changes:notebooks/multiome-archetypes MAGIC imputed.Rmd

# Visualize what headings are called so that you can extract them to form a dataframe
Seurat::Embeddings(object = sclc, reduction = "umap")

sclc[["cluster"]] <- Seurat::Idents(object = sclc)
```
```{r}
DimHeatmap(TKO, dims = 1:20, cells = 500, balanced = TRUE)
# TKO <- JackStraw(TKO, num.replicate = 100)
# TKO <- ScoreJackStraw(TKO, dims = 1:20)
```


```{r, echo = FALSE}
# Prepare a dataframe for cell plotting
plot.data <- Seurat::FetchData(object = sclc, vars = c("UMAP_1", "UMAP_2", "cluster"))

# Make a column of row name identities (these will be your cell/barcode names)
plot.data$label <- paste(rownames(plot.data))
# install.packages('plotly')


# Plot your data, in this example my Seurat object had 21 clusters (0-20)
plot_ly(data = plot.data, 
        x = ~UMAP_1, y = ~UMAP_2,
        color = ~cluster,
        mode = "markers", 
        colors = palette(rainbow(20)),
        type = 'scatter',
        marker = list(size = 5, width=2), # controls size of points
        text=~label, #This is that extra column we made earlier for which we will use for cell ID
        hoverinfo="text") #When you visualize your plotly object, hovering your mouse pointer over a point shows cell names
```

# Plot PCA of archetypes
```{r, include = FALSE}
# Extract tSNE information from Seurat Object
pca_1 <- sclc[["pca"]]@cell.embeddings[,1]
pca_2 <- sclc[["pca"]]@cell.embeddings[,2]
pca_3 <- sclc[["pca"]]@cell.embeddings[,3]


# Visualize what headings are called so that you can extract them to form a dataframe
Seurat::Embeddings(object = sclc, reduction = "pca")
```



```{r, echo = FALSE}
plot.data <- Seurat::FetchData(object = sclc, vars = c("PC_1", "PC_2","PC_3", "cluster"))
plot.data$label <- paste(rownames(plot.data))
plot_ly(data = plot.data, 
        x = ~PC_1, y = ~PC_2,z = ~PC_3,
        color = ~cluster,
        mode = "markers", 
        colors = palette(rainbow(20)),
        type = 'scatter3d',
        marker = list(size = 5, width=2), # controls size of points
        text=~label, #This is that extra column we made earlier for which we will use for cell ID
        hoverinfo="text") #
# VizDimLoadings(data, dims = 1:3, reduction = "pca")
```
<<<<<<< Updated upstream:archetype-plasticity-notebooks/M1/multiome-archetypes.Rmd

# Run ParetoTI on SCTransformed scaled.data (via the PCA projection from Debbie's analysis)
```{r paretoti, message=FALSE}

x <- sclc@assays$SCT@scale.data
x_pca <- read.csv('../data/data_from_debbie/pca-embedding.csv',header = TRUE,row.names = 1)
x_pca <- t(x_pca)
x_pca <- x_pca[1:20,] #keep only top 18 PCs
loadings<- read.csv('../data/data_from_debbie/pca-feature-loadings-projected.csv', header = TRUE, row.names = 1)
loadings <- as.matrix(loadings)
##################################
# load("../data/arc.Robj", verbose=TRUE) #from code commented out below
# load("../data/arc_ks.Robj", verbose=TRUE)

arc_ks = ParetoTI::k_fit_pch(x_pca, ks = 2:8, check_installed = T,
                   bootstrap = T, bootstrap_N = 100, maxiter = 1000,
                   bootstrap_type = "m", seed = 2543,
                   volume_ratio = "t_ratio", # set to "none" if too slow
                   delta=0, conv_crit = 1e-04, order_type = "align",
                   sample_prop = 0.75)
```

```{r}
# # Show variance explained by a polytope with each k (cumulative)
ParetoTI::plot_arc_var(arc_ks, type = "varexpl", point_size = 2, line_size = 1.5) + theme_bw()
# Show variance explained by k-vertex model on top of k-1 model (each k separately)
ParetoTI::plot_arc_var(arc_ks, type = "res_varexpl", point_size = 2, line_size = 1.5) + theme_bw()
# Show variance in position of vertices obtained using bootstraping 
# - use this to find largest k that has low variance
ParetoTI::plot_arc_var(arc_ks, type = "total_var", point_size = 2, line_size = 1.5) +
  theme_bw() +
  ylab("Mean variance in position of vertices")
# Show t-ratio
ParetoTI::plot_arc_var(arc_ks, type = "t_ratio", point_size = 2, line_size = 1.5) + theme_bw()
```
I believe the mean variance of arc 1 is really high because this is the ciliated cells, and there are only a few of them so bootstrapping causes the position to change a lot.

```{r arc_rob}
# Fit # of archetypes from above plot
arc <- fit_pch((x_pca), noc = 4)
# Fit 5 archetypes with bootstrapping for robustness
arc_rob = fit_pch_bootstrap(x_pca, n = 200, sample_prop = .8, seed = 2543, delta = 1,
                            noc = 4)
arc_ave <- average_pch_fits(arc_rob)

save(arc, file="../data/data_from_debbie/arc.Robj")
save(arc_ks, file="../data/data_from_debbie/arc_ks.Robj")
save(arc_rob, file="../data/data_from_debbie/arc_rob.Robj")
save(arc_ave, file="../data/data_from_debbie/arc_ave.Robj")


```

```{r, include = FALSE}


# cols <- c(brewer.pal(9, "Set1"),'gray')

ParetoTI::plot_arc(arc_data = arc, data = x_pca,
                   which_dimensions = 1:2,colors = palette(rainbow(20)),
                  data_lab = as.character(Seurat::Idents(sclc))) + theme_bw()

p_pca = plot_arc(arc_data = arc, data = x_pca, 
                 which_dimensions = 1:3, line_size = 1.5,
                  data_lab = as.character(Seurat::Idents(sclc)),
                 colors = palette(rainbow(20)),
                 text_size = 60, data_size = 4) 
plotly::layout(p_pca, title = "Average Archetypes for Top 20 PCs")
# htmlwidgets::saveWidget(p_pca, "archetypes.html")


```
## Without ciliated cells

It looks like ciliated cells are dominating PC2, so we are going to remove those cells and just look at the four archetypes defining the other cell types. 
```{r without-ciliated, message=FALSE}
sclc_small <- subset(x = sclc, subset = cluster != "Cilliated cells")

DefaultAssay(sclc_small) <- "RNA"
# store mitochondrial percentage in object meta data
sclc_small <- PercentageFeatureSet(sclc_small, pattern = "mt-", col.name = "percent.mt")

# run sctransform
# BiocManager::install("glmGamPoi")
library(glmGamPoi)
sclc_small <- SCTransform(sclc_small, vars.to.regress = "percent.mt", verbose = FALSE,method = "glmGamPoi")

# These are now standard steps in the Seurat workflow for visualization and clustering

sclc_small <- Seurat::ScaleData(object = sclc_small)

sclc_small <- Seurat::RunPCA(object = sclc_small)
sclc_small <- Seurat::ProjectDim(sclc_small, reduction = 'pca')

x_pca_small <- sclc_small@reductions$pca@cell.embeddings
x_pca_small <- t(x_pca_small)
x_pca_small <- x_pca_small[1:20,] #keep only top 18 PCs
loadings<- sclc_small@reductions$pca@feature.loadings.projected
loadings <- as.matrix(loadings)
write.csv(sclc_small@assays$SCT@scale.data, '../data/data_from_debbie/scaled-SCT-data_small.csv')
write.csv(sclc_small@reductions$pca@cell.embeddings, '../data/data_from_debbie/pca-embedding_small.csv')
write.csv(sclc_small@reductions$pca@feature.loadings, '../data/data_from_debbie/pca-feature-loadings_small.csv')

write.csv(sclc@reductions$pca@feature.loadings.projected, '../data/data_from_debbie/pca-feature-loadings-projected_small.csv')
save(sclc_small, file = '../data/data_from_debbie/sclc-no-ciliated-cells.Robj')
write.csv(sclc_small@reductions$umap@cell.embeddings, '../data/data_from_debbie/umap-embedding_small.csv')

```


## Plot PCA of smaller dataset (no ciliated cells)
```{r, include = FALSE}
# load(file = '../data/data_from_debbie/sclc-no-ciliated-cells.Robj')

# Extract tSNE information from Seurat Object
pca_1 <- sclc_small[["pca"]]@cell.embeddings[,1]
pca_2 <- sclc_small[["pca"]]@cell.embeddings[,2]
pca_3 <- sclc_small[["pca"]]@cell.embeddings[,3]


# Visualize what headings are called so that you can extract them to form a dataframe
Seurat::Embeddings(object = sclc_small, reduction = "pca")
```



```{r, echo = FALSE}
plot.data <- Seurat::FetchData(object = sclc_small, vars = c("PC_1", "PC_2","PC_3", "cluster"))
plot.data$label <- paste(rownames(plot.data))
plot_ly(data = plot.data, 
        x = ~PC_1, y = ~PC_2,z = ~PC_3,
        color = ~cluster,
=======
# Labeling cells
Somehow we lost the cell labels, so we'll read those in from Debbie and use them to plot. 

```{r}
labels <- read.table('../data/meta_cells.txt', row.names = 1)
TKO[['new.ident']] = labels
Idents(TKO) <- TKO[['new.ident']]
plot.data <- Seurat::FetchData(object = TKO, vars = c("PC_1", "PC_2","PC_3", "new.ident"))
plot.data$label <- paste(rownames(plot.data))
saveRDS(object = TKO, file = "../data/TKO_MAGIC_updated.rds")

plot_ly(data = plot.data, 
        x = ~PC_1, y = ~PC_2,z = ~PC_3,
        color = ~new.ident,
>>>>>>> Stashed changes:notebooks/multiome-archetypes MAGIC imputed.Rmd
        mode = "markers", 
        colors = palette(rainbow(20)),
        type = 'scatter3d',
        marker = list(size = 5, width=2), # controls size of points
        text=~label, #This is that extra column we made earlier for which we will use for cell ID
        hoverinfo="text") #
# VizDimLoadings(data, dims = 1:3, reduction = "pca")
<<<<<<< Updated upstream:archetype-plasticity-notebooks/M1/multiome-archetypes.Rmd
```

``` {r}
=======

```



# Run archetype analysis on imputed data
``` {r}

TKO <- readRDS( file = "../data/TKO_MAGIC_updated.rds")
TKO_pca <- TKO@reductions$pca@cell.embeddings
TKO_pca <- t(TKO_pca)
TKO_pca <- TKO_pca[1:20,] #keep only top 18 PCs
loadings<- TKO@reductions$pca@feature.loadings.projected
loadings <- as.matrix(loadings)

>>>>>>> Stashed changes:notebooks/multiome-archetypes MAGIC imputed.Rmd
##################################

arc_ks = ParetoTI::k_fit_pch(TKO_pca, ks = 2:8, check_installed = T,
                   bootstrap = T, bootstrap_N = 200, maxiter = 1000,
                   bootstrap_type = "m", seed = 2543,
                   volume_ratio = "t_ratio", # set to "none" if too slow
                   delta=0, conv_crit = 1e-04, order_type = "align",
                   sample_prop = 0.75)

# # Show variance explained by a polytope with each k (cumulative)
ParetoTI::plot_arc_var(arc_ks, type = "varexpl", point_size = 2, line_size = 1.5) + theme_bw()
# Show variance explained by k-vertex model on top of k-1 model (each k separately)
ParetoTI::plot_arc_var(arc_ks, type = "res_varexpl", point_size = 2, line_size = 1.5) + theme_bw()
# Show variance in position of vertices obtained using bootstraping 
# - use this to find largest k that has low variance
ParetoTI::plot_arc_var(arc_ks, type = "total_var", point_size = 2, line_size = 1.5) +
  theme_bw() +
  ylab("Mean variance in position of vertices")
# Show t-ratio
ParetoTI::plot_arc_var(arc_ks, type = "t_ratio", point_size = 2, line_size = 1.5) + theme_bw()

# 
```


```{r}
# Fit 4 archetypes
arc <- fit_pch((TKO_pca), noc = 4)
# Fit 4 archetypes with bootstrapping for robustness
arc_rob = fit_pch_bootstrap(TKO_pca, n = 200, sample_prop = .8, seed = 2543, delta = 1,
                            noc = 4)
arc_ave <- average_pch_fits(arc_rob)


p_pca = plot_arc(arc_data = arc, data = TKO_pca, 
                 which_dimensions = 1:3, line_size = 1.5,
                  data_lab = (Seurat::Idents(TKO)),
                 colors = palette(rainbow(20)), mode = "markers", 
                 text_size = 60, data_size = 4) 
plotly::layout(p_pca, title = "Average Archetypes for Top 20 PCs")
```
```{r}
arc_tsne = arch_to_tsne(arc, TKO_pca, which_dimensions = 1:2)
plot_arc(arc_data = arc_tsne$arc_data, data = arc_tsne$data,data_lab = (Seurat::Idents(TKO)),
                 colors = palette(rainbow(20)),
    which_dimensions = 1:2) +
    theme_bw()
# install.packages(
# 'umap'
# )
arc_umap = arch_to_umap(arc, TKO_pca, which_dimensions = 1:2,
                        method = c( 'naive'))# implemented in R and slow
                                   #"umap-learn")) # requires python module
plot_arc(arc_data = arc_umap$arc_data, data = arc_umap$data,data_lab = (Seurat::Idents(TKO)),
                 colors = palette(rainbow(20)),
    which_dimensions = 1:2) +
    theme_bw()
```

```{r}
p_pca = plot_arc(arc_data = arc, data = x_pca_small, 
                 which_dimensions = 1:3, line_size = 1.5,
                 data_lab = sclc_small@assays$SCT@data['Scgb1a1',],
                 text_size = 60, data_size = 4) 
plotly::layout(p_pca, title = "Average Archetypes for Top 20 PCs")
```

```{r}

save(arc, file="../data/arc_updated.Robj")
save(arc_ks, file="../data/arc_ks_updated.Robj")
save(arc_rob, file="../data/arc_rob_updated.Robj")
save(arc_ave, file="../data/arc_ave_updated.Robj")
```

# Randomize data to compute background distirbution of scores

This method is used to compare the archetypes on the data to randomized data-- how likely is it to explain a similar amount of variance, get a similar t-ratio, or similar variance in archetype position, by chance?

94% of the variance (proportion of the top 18 PC's explained variance, not total variance in dataset) is explained by the archetypes (5 dimensions). This has a p-value of 0.001. 

```{r}
# start = Sys.time()
# pch_rand = randomise_fit_pch(x_pca, arc_data = arc,
#                              n_rand = 1000,
#                              replace = FALSE, 
#                              volume_ratio = "t_ratio",
#                              maxiter = 500, delta = 1, conv_crit = 1e-4)
# # use type m to run on a single machine or cloud
# # type = "m", clust_options = list(cores = 3))
# # use clustermq (type cmq) to run as jobs on a computing cluster (higher parallelisation)
# # type = "cmq", clust_options = list(njobs = 10)) 
# 
# # This analysis took:
# Sys.time() - start
# plot(pch_rand, type = c("t_ratio"), nudge_y = 5)

```

# Determine enriched genes and gene sets to define archetypes

This will be especially interesting for comparing archetypes 4 and 5, since it is pretty clear that the shape of the data is not clearly defined without an archetype at 5, but not many cells actually lie close to it. We use the ParetoTI package to evaulate gene sets enriched in each location. 

```{r}
library(matrixStats)

# x.orig <- as.matrix(data[['RNA']]@data)
# 
# rownames(x.orig)<-gsub("-", ".", rownames(x.orig))
# 
# 
# 
# ix <- which(rownames(x.orig) %in%c('1.Sep','10.Sep','11.Mar','11.Sep','2.Sep','3.Mar','3.Sep','4.Mar','4.Sep','5.Sep', '5.Mar','6.Sep','6.Mar','7.Sep','7.Mar','8.Sep','8.Mar','9.Sep','9.Mar','RP11-206L10.1'))
# ix <- which(rownames(x.orig) %in% c('7SK.1'))
# clean <- x.orig[-ix, ]

# data <- read.csv('../data/data_from_debbie/scaled-SCT-data_small.csv', header = TRUE,row.names = 1)
data <- as.matrix(TKO@assays$Imputed_counts@data)
activ = measure_activity(as.matrix(data), activity_method = 'pseudoinverse',# row names are assumed to be gene identifiers,
                         which = 'BP', return_as_matrix = F,
                         taxonomy_id = 10090, keytype = "ALIAS", #10090 is mouse
                         lower = 10, upper = 1000)
                         # aucell_options =list(aucMaxRank =
                          # nrow(as.matrix(x.orig)) * 0.05, binary = F, nCores = 3, plotStats = TRUE))
save(activ, file="../data/data_from_debbie/activ_imputed.Robj")
# load("../data/data_from_debbie/activ_pi.Robj")
# activ_pi <- within(activ_pi, rm('2__deoxyribonucleotide_biosynthetic_process','2__deoxyribonucleotide_metabolic_process','2_oxoglutarate_metabolic_process','3__phosphoadenosine_5__phosphosulfate_metabolic_process',
# '3__UTR_mediated_mRNA_destabilization',
# '3__UTR_mediated_mRNA_stabilization',
# '7_methylguanosine_mRNA_capping',
# '7_methylguanosine_RNA_capping',
# '4_hydroxyproline_metabolic_process'))
# 
# activ_pi <- within(activ_pi, rm(`_de_novo__posttranslational_protein_folding`,
# `_de_novo__protein_folding`,
# `poly_A_+_mRNA_export_from_nucleus`))


data_attr = merge_arch_dist(arc_data = arc, data = TKO_pca, 
                            feature_data = as.matrix(data),
                            colData = activ, 
                            dist_metric = c("euclidean", "arch_weights")[1],
                            colData_id = "cells", rank = F) 

enriched_genes = find_decreasing_wilcox(data_attr$data, data_attr$arc_col,
                                features = data_attr$features_col,
                                bin_prop = 0.05, method = "BioQC")
write.csv(enriched_genes, '../data/data_from_debbie/enriched-genes.csv')

enriched_sets = find_decreasing_wilcox(data_attr$data, data_attr$arc_col,
                                features = data_attr$colData_col,
                                bin_prop = 0.05, method = "BioQC")

labs = get_top_decreasing(summary_genes = enriched_genes, summary_sets = enriched_sets,
                          cutoff_genes = 0.05,cutoff_sets = 0.05,
                          cutoff_metric = "wilcoxon_p_val", 
                          p.adjust.method = "fdr", 
                          order_by = "mean_diff", order_decreasing = T)

p_pca = plot_arc(arc_data = arc, data = x_pca_small,
                 which_dimensions = 1:3, line_size = 1.5,
                 data_lab = activ_pi$neural_crest_cell_development,
                 text_size = 60, data_size = 6)
plotly::layout(p_pca, title = "neural crest activity")


#below takes forever
# enriched_genes_gam = find_decreasing(data_attr$data, data_attr$arc_col,
#                                 features = data_attr$features_col, return_only_summary = TRUE)
# write.csv(enriched_genes_gam, '../data/single-cell/enriched_genes_gam.csv')
# 
# enriched_sets_gam = find_decreasing(data_attr$data, data_attr$arc_col,
#                                 features = data_attr$colData_col, return_only_summary = TRUE)
# write.csv(enriched_sets_gam, '../data/single-cell/enriched_sets_gam.csv')
# 
# 
# 
# labs_gam = get_top_decreasing(summary_genes = enriched_genes_gam, summary_sets = enriched_sets_gam,
#                           cutoff_genes = 0.05,cutoff_sets = 0.05,
#                           cutoff_metric = "mean_prob", 
#                           p.adjust.method = "none", 
#                           order_by = "deriv50", order_decreasing = F,
#                           min_max_diff_cutoff_g = .05)
# 
# fit_arc_gam_1('ASCL1', "archetype_2", data_attr)
```

# Only NE populations

```{r}

NE <- subset(x = TKO, idents = c( "NFIB high_NE",  "NE_1", "NE_2", "NE_3","NE_4","NE_6",  "NE_5",  "NE_7","NE_8"))

DefaultAssay(NE) <- "Imputed_counts"

NE <- Seurat::ScaleData(object = NE)
NE<- Seurat::FindVariableFeatures(NE)
NE <- Seurat::RunPCA(object = NE)
NE <- Seurat::ProjectDim(NE, reduction = 'pca')

NE_pca <- NE@reductions$pca@cell.embeddings
NE_pca <- t(NE_pca)
ElbowPlot(NE)

NE_pca <- NE_pca[1:20,] #keep only top 18 PCs
loadings<- NE@reductions$pca@feature.loadings.projected
loadings <- as.matrix(loadings)
NE[["cluster"]] = Idents(NE)
plot.data <- Seurat::FetchData(object = NE, vars = c("PC_1", "PC_2","PC_3", "cluster"))
plot.data$label <- paste(rownames(plot.data))
plot_ly(data = plot.data, 
        x = ~PC_1, y = ~PC_2,z = ~PC_3,
        color = ~cluster,
        mode = "markers", 
        colors = palette(rainbow(20)),
        type = 'scatter3d',
        marker = list(size = 5, width=2), # controls size of points
        text=~label, #This is that extra column we made earlier for which we will use for cell ID
        hoverinfo="text") #

```

```{r}
arc_ne = ParetoTI::k_fit_pch(NE_pca, ks = 2:8, check_installed = T,
                   bootstrap = T, bootstrap_N = 200, maxiter = 1000,
                   bootstrap_type = "m", seed = 2543,
                   volume_ratio = "t_ratio", # set to "none" if too slow
                   delta=0, conv_crit = 1e-04, order_type = "align",
                   sample_prop = 0.75)

# # Show variance explained by a polytope with each k (cumulative)
ParetoTI::plot_arc_var(arc_ne, type = "varexpl", point_size = 2, line_size = 1.5) + theme_bw()
# Show variance explained by k-vertex model on top of k-1 model (each k separately)
ParetoTI::plot_arc_var(arc_ne, type = "res_varexpl", point_size = 2, line_size = 1.5) + theme_bw()
# Show variance in position of vertices obtained using bootstraping 
# - use this to find largest k that has low variance
ParetoTI::plot_arc_var(arc_ne, type = "total_var", point_size = 2, line_size = 1.5) +
  theme_bw() +
  ylab("Mean variance in position of vertices")
# Show t-ratio
ParetoTI::plot_arc_var(arc_ne, type = "t_ratio", point_size = 2, line_size = 1.5) + theme_bw()


```
```{r}

# Fit 4 archetypes
arc <- fit_pch((NE_pca), noc = 4)
# Fit 4 archetypes with bootstrapping for robustness
arc_rob = fit_pch_bootstrap(NE_pca, n = 200, sample_prop = .8, seed = 2543, delta = 1,
                            noc = 4)
arc_ave <- average_pch_fits(arc_rob)


p_pca = plot_arc(arc_data = arc_ave, data = NE_pca, 
                 which_dimensions = 1:3, line_size = 1.5,
                  data_lab = (Seurat::Idents(NE)),
                 colors = palette(rainbow(20)),
                 text_size = 60, data_size = 4) 
plotly::layout(p_pca, title = "Average Archetypes for Top 20 PCs")
```

```{r}
library(matrixStats)

# x.orig <- as.matrix(data[['RNA']]@data)
# 
# rownames(x.orig)<-gsub("-", ".", rownames(x.orig))
# 
# 
# 
# ix <- which(rownames(x.orig) %in%c('1.Sep','10.Sep','11.Mar','11.Sep','2.Sep','3.Mar','3.Sep','4.Mar','4.Sep','5.Sep', '5.Mar','6.Sep','6.Mar','7.Sep','7.Mar','8.Sep','8.Mar','9.Sep','9.Mar','RP11-206L10.1'))
# ix <- which(rownames(x.orig) %in% c('7SK.1'))
# clean <- x.orig[-ix, ]

# data <- read.csv('../data/data_from_debbie/scaled-SCT-data_small.csv', header = TRUE,row.names = 1)
data <- as.matrix(NE@assays$Imputed_counts@data)
activ = measure_activity(as.matrix(data), activity_method = 'pseudoinverse',# row names are assumed to be gene identifiers,
                         which = 'BP', return_as_matrix = F,
                         taxonomy_id = 10090, keytype = "ALIAS", #10090 is mouse
                         lower = 10, upper = 1000)
                         # aucell_options =list(aucMaxRank =
                          # nrow(as.matrix(x.orig)) * 0.05, binary = F, nCores = 3, plotStats = TRUE))
save(activ, file="../data/activ_imputed_NE.Robj")

###########################

load("../data/activ_imputed_NE.Robj")
# activ_pi <- within(activ_pi, rm('2__deoxyribonucleotide_biosynthetic_process','2__deoxyribonucleotide_metabolic_process','2_oxoglutarate_metabolic_process','3__phosphoadenosine_5__phosphosulfate_metabolic_process',
# '3__UTR_mediated_mRNA_destabilization',
# '3__UTR_mediated_mRNA_stabilization',
# '7_methylguanosine_mRNA_capping',
# '7_methylguanosine_RNA_capping',
# '4_hydroxyproline_metabolic_process'))
# 
# activ_pi <- within(activ_pi, rm(`_de_novo__posttranslational_protein_folding`,
# `_de_novo__protein_folding`,
# `poly_A_+_mRNA_export_from_nucleus`))


data_attr = merge_arch_dist(arc_data = arc_ave, data = NE_pca, 
                            feature_data = as.matrix(data),
                            colData = activ, 
                            dist_metric = c("euclidean", "arch_weights")[1],
                            colData_id = "cells", rank = T) 
ans <- data_attr$data[,1:5]
save(arc, file="../data/arc_NE.Robj")
save(arc_rob, file="../data/arc_rob_NE.Robj")
save(arc_ave, file="../data/arc_ave_NE.Robj")
save(data_attr, file="../data/data_attr_NE.Robj")
write.csv(ans, '../data/arc_distance_inputed-ne.csv')

```
At this point, we saved the data as a csv with distances to each archetype and all the expression data. We'l read this into python and make labels for each cell for each archetype: "10_pct_closest_arc1", "10_pct_closest_arc2", etc.

```{r}

labels <- read.table('../data/labels_arc_dist.csv', row.names = 1, sep = ',', header = TRUE)

NE[['arc.distance.ident']] = labels
Idents(NE) <- NE[['arc.distance.ident']]
saveRDS(object = NE, file = "../data/NE_MAGIC_updated.rds")

for_cibersort <- subset(x = NE, idents = c( "archetype_1_closest",  "archetype_2_closest", "archetype_3_closest", "archetype_4_closest"))
cibersort_data <- as.matrix(for_cibersort@assays$Imputed_counts@data)
colnames(cibersort_data) <- Idents(for_cibersort)
rownames(cibersort_data)<-toupper(rownames(cibersort_data))
write.csv(cibersort_data, '../data/cibersort-ne-cells-closest-to-archetypes.csv')
write.table(cibersort_data, '../data/cibersort-ne-cells-closest-to-archetypes.txt', sep = "\t")

saveRDS(object = for_cibersort, file = "../data/NE_ForCibersort_MAGIC_updated.rds")

cibersort_data_clustered <- as.matrix(NE@assays$Imputed_counts@data)
Idents(NE) <- NE[['new.ident']]
colnames(cibersort_data_clustered) <- Idents(NE)
rownames(cibersort_data_clustered)<-toupper(rownames(cibersort_data_clustered))
write.table(cibersort_data_clustered, '../data/cibersort-ne-cells-clustered.txt', sep = "\t")
write.table(Idents(NE), '../data/NE_cell_clusters.txt', sep = "\t")


```


``` {r pca-plots}

p_pca = plot_arc(arc_data = arc_ave, data = NE_pca, 
                 which_dimensions = 1:3, line_size = 1.5,
                  data_lab = (Seurat::Idents(NE)),
                 colors = palette(rainbow(20)),
                 text_size = 60, data_size = 4) 
plotly::layout(p_pca, title = "Average Archetypes for Top 20 PCs")


p_pca = plot_arc(arc_data = arc_ave, data = NE_pca, 
                 which_dimensions = 1:3, line_size = 1.5,
                  data_lab = as.numeric(data_attr$data$archetype_4),
                 text_size = 60, data_size = 4) 
plotly::layout(p_pca, title = "Average Archetypes for Top 20 PCs")
install.packages('htmlwidgets')
library(htmlwidgets)

htmlwidgets::saveWidget(p_pca, "closest_to_arc.html")

```



```{r enrichments}
enriched_genes = find_decreasing_wilcox(data_attr$data, data_attr$arc_col,
                                features = data_attr$features_col,
                                bin_prop = 0.05, method = "BioQC")
write.csv(enriched_genes, '../data/enriched-genes-imputed-ne.csv')

enriched_sets = find_decreasing_wilcox(data_attr$data, data_attr$arc_col,
                                features = data_attr$colData_col,
                                bin_prop = 0.05, method = "BioQC")
write.csv(enriched_sets, '../data/enriched-sets-imputed-ne.csv')

labs = get_top_decreasing(summary_genes = enriched_genes, summary_sets = enriched_sets,
                          cutoff_genes = 0.05,cutoff_sets = 0.05,
                          cutoff_metric = "wilcoxon_p_val", 
                          p.adjust.method = "fdr", 
                          order_by = "mean_diff", order_decreasing = T)

p_pca = plot_arc(arc_data = arc, data = x_pca_small,
                 which_dimensions = 1:3, line_size = 1.5,
                 data_lab = activ_pi$neural_crest_cell_development,
                 text_size = 60, data_size = 6)
plotly::layout(p_pca, title = "neural crest activity")

```

